The Gillespie Algorithm: Stochastic Modeling of Molecular Reaction Networks

### 2024-11-27 paper.Rmd
update

bibliography: references.bib 
editor_options: 
  markdown: 
    wrap: 72
---

\newpage 
\tableofcontents 
\newpage

@@

Combining the two factors, we obtain the master equation with the form $^{[1]}$ $$\frac{\partial}{\partial t}P(S,t)=\sum_{S'}P(S',t)R(S',S)-\sum_{S'}P(S,t)R(S,S')$$, where $R(S,S')$ represents the reaction rate from the state $S$ to the state $S'$.

@@

In the article `Cellular growth and division in the Gillespie algorithm` $^{[2]}$, the authors discussed the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

For example, in a second order reaction $A+B\rightarrow AB$, we have $$\frac{d}{dt}X_{AB}=V\frac{d}{dt}[AB]=V*c'[A][B]=\frac{c'}{V}*X_AX_B=c*h(Y)$$, where $c'$ is the true reaction constant measured in chemistry, $V$ is the volume of the system and $[i]=\frac{X_i}{V}$ is the number concentration of species $i$.

With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$, here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.

@@

A practical alternative to solve the master equation is the stochastic simulation approach devised by Gillespie $^{[3]}$. Using the idea that $P(Y,t|Y_0,t_0)$ describes the ensemble average of all possible micro-states given initial state $Y_0$ at time $t_0$, the Gillespie algorithm generates an ensemble of sample trajectories of the system with statistics which asymptotically converges to the solution of the corresponding Master Equation.

@@

Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP.

@@

```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (1 - (37 + 2) / 1e6) * 10   # No dimerization occurs
c2 <- (37 / 1e6) * 10             # Formation of CPD
c3 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T2"],               # Reaction 1
    c2 * state["T1"] * state["T2"], # Reaction 2
    c3 * state["T2"]                # Reaction 3
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

@@

    # Update state based on reaction
    if (reaction == 1) state["T2"] <- state["T2"] - 1
    if (reaction == 2) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["CPD"] <- state["CPD"] + 1
    }
    if (reaction == 3) {
      state["T2"] <- state["T2"] - 1
      state["64PP"] <- state["64PP"] + 1
    }

@@

```{r graph_1, echo=FALSE, warning=FALSE, message=FALSE}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")
# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "Decay-Dimerization Reaction Set Simulation",
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()
```

@@

```{r decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}

@@

```{r decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
# Logarithmic-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (log(R))"
  ) +
  theme_minimal()

@@

[1]. N. G. V. Kampen, Stochastic Processes in Physics and Chemistry. North-Holland, Amsterdam, 1992.

[2]. T. Lu, D. Volfson, L. Tsimring, J. Hasty, Cellular growth and division in the Gillespie algorithm, Syst Biol (Stevenage). 2004 Jun;1(1):121-8

[3]. D. T. Gillespie, “Exact stochastic simulation of coupled chemical reactions,” J. Phys. Chem., vol. 81, no. 25, pp. 2340–2361,
1977.

>>>>

bibliography: references.bib 
editor_options: 
  markdown: 
    wrap: 72
---
\newpage
\tableofcontents
\newpage

@@

Combining the two factors, we obtain the master equation with the form (@vanKampen1992) $$\frac{\partial}{\partial t}P(S,t)=\sum_{S'}P(S',t)R(S',S)-\sum_{S'}P(S,t)R(S,S')$$ where $R(S,S')$ represents the reaction rate from the state $S$ to the state $S'$.

@@

In the article `Cellular growth and division in the Gillespie algorithm` (@Lu2004), the authors discussed the system of well-stirred mixture of $N$ chemical species. Since the system is well-stirred, the distribution of particles is always uniform in space. The state of the system can be described using a N-vector $Y=(X_1,...,X_N)$, where $X_i$ is the number of molecules of type $i$ in the system. Assume there are $\mu=1,...,M$ elementary reaction channels $R_{\mu}$. Let $c_{\mu}$ be the reaction rate, i.e., the probability that a random combination of molecules from channel $R_{\mu}$ selected at the moment $t$ react in the interval $[t,t+dt)$ with probability $c_{\mu}dt+o(dt)$. Let $h_{\mu}(Y)$ be the total number of possible distinct combinations of molecules for a channel $R_{\mu}$ when the system is in state $Y$, and $\alpha_{\mu}=(\alpha_{1,\mu},...,\alpha_{N,\mu})$ is a constant stoichiometric vector prescribing the change in the state of the system after the reaction $R_{\mu}$ has occurred.

For example, in a second order reaction $A+B\rightarrow AB$, we have $$\frac{d}{dt}X_{AB}=V\frac{d}{dt}[AB]=V*c'[A][B]=\frac{c'}{V}*X_AX_B=c*h(Y)$$ where $c'$ is the true reaction constant measured in chemistry, $V$ is the volume of the system and $[i]=\frac{X_i}{V}$ is the number concentration of species $i$.

With the above setting-up, The master equation for this system becomes $$\frac{\partial}{\partial t}P(Y,t|Y_0,t_0)=\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y-\alpha_{\mu})P(Y-\alpha_{\mu},t|Y_0,t_0)-\sum_{\mu=1}^M c_{\mu}h_{\mu}(Y)P(Y,t|Y_0,t_0)$$ here $P(Y,t|Y_0,t_0)$ is the probability of the system being in state $Y$ at time $t$ given it is in state $Y_0$ at time $t_0$.

@@

A practical alternative to solve the master equation is the stochastic simulation approach devised by Gillespie (@Gillespie1977). Using the idea that $P(Y,t|Y_0,t_0)$ describes the ensemble average of all possible micro-states given initial state $Y_0$ at time $t_0$, the Gillespie algorithm generates an ensemble of sample trajectories of the system with statistics which asymptotically converges to the solution of the corresponding Master Equation.

@@

Therefore, this simulation involves 4 species and 3 reactions channels: pyrimidine monomers Thymine 1 (T1), Thymine 2 (T2), and dimer photoproducts CPD and 64-PP. 
\[
T1 + T2 \xrightarrow{c1} CPD
\]
\[
T1 + T2 \xrightarrow{c2} 64PP
\]

@@

```{r simulation_1, echo=FALSE, warning=FALSE, message=FALSE}
# Load required libraries
if (!require("reshape2")) install.packages("reshape2")
if (!require("ggplot2")) install.packages("ggplot2")
library(reshape2)
library(ggplot2)
# Define reaction rates
c1 <- (37 / 1e6) * 10             # Formation of CPD
c2 <- (2 / 1e6) * 10              # Formation of 64-PP
# Initial state
state <- c(T1 = 10000, T2 = 10000, CPD = 0, `64PP` = 0)
# Define propensity functions
propensities <- function(state) {
  c(
    c1 * state["T1"] * state["T2"], # Reaction 1
    c2 * state["T1"] * state["T2"]  # Reaction 2
  )
}
# Gillespie simulation function
simulate_gillespie <- function(state, max_time) {
  # Initialize time and output data frame
  time <- 0
  output <- data.frame(
    Time = numeric(0),
    T1 = numeric(0),
    T2 = numeric(0),
    CPD = numeric(0),
    `64PP` = numeric(0)
  )

@@

    # Update state based on reaction
    if (reaction == 1) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["CPD"] <- state["CPD"] + 1
    }
    if (reaction == 2) {
      state["T1"] <- state["T1"] - 1
      state["T2"] <- state["T2"] - 1
      state["64PP"] <- state["64PP"] + 1
    }

@@

```{r fig-dimerization-sim, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Thymine Photodimerization Simulation"}
# Reshape data for plotting
simulation_long <- melt(simulation, id.vars = "Time")
# Plot results
ggplot(simulation_long, aes(x = Time, y = value, color = variable)) +
  geom_line() +
  labs(
    title = "Dimerization Reaction Set Simulation",
    x = "Time",
    y = "Molecule Count",
    color = "Species"
  ) +
  theme_minimal()
```

Both CPD and 64-PP formation follow logarithmic growth, as expected under specific UV radiation according to previously reported quantum yields (@MartinezFernandez2022; @Mouret2006). The number of reactions per unit time plateaus as the pyrimidine pairs decrease, consistent with the propensity functions.
Due to the stochastic nature of molecular reactions, this algorithm reflects fluctuations of photodimerization as it models the probabilistic process of each reaction.

### Biological Significance
- **Dimerization**
As pyrimidine photodimerization transforms two pyrimidine monomers into a dimer, this simulation has accurately represented the formation of two possible dimers, CPD and 64-PP, from pairs of thymine monomers. Dimerization is also known to occur between pairs of cytosine, as well as between thymine and cytosine, with varying rates based on the monomers involved as well as the level of UV radiation (@Mouret2006).
- **Real-World Application**
Dimerization is widely studied, as it can lead to DNA mutations causing skin cancer.The Gillespie  algorithm allows us to realistically model these reactions, as it allows us to demonstrate that our parameters of interest change over time.
This algorithm can be further adjusted to show molecular reactions with various molecular species and reaction channels, with applications in biology, chemistry, and bioinformatics.

@@

```{r fig-decay-linear, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Linear Scale"}

@@

```{r fig-decay-log, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Radioactive Decay Simulation in Logarithmic Scale"}
# Logarithmic-scale plot
ggplot(simulation, aes(x = Time, y = R)) +
  geom_line(color = "red") +
  scale_y_log10() +
  labs(
    x = "Time",
    y = "Remaining Radioactive Substance (log(R))"
  ) +
  theme_minimal()

@@
